# Your First SPC Charts With Base R {#first-chart}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/3,
                      dev     = 'svg')
```

From part 1 of this book we have a good grasp of what SPC is and how SPC charts work. In this chapter we will start constructing SPC chart. We begin using only functions from base R. Later we will move to `ggplot2`and `qicharts2`.

In essence, an SPC charts is a (point-and-)line plot of data over time with a horizontal line to represent the data centre and -- in case of a control chart -- two lines to represent the estimated upper and lower bounds of the natural variation in data.

To construct an SPC chart, all we need are four vectors with the y-coordinates of the data points and lines. 

Let's start by creating some artificial data to work with. We draw 24 random numbers from a normal distribution with mean = 0 and standard deviation (sd) = 1. 

```{r}
# Initialise random number generator to make results reproducible
set.seed(27)

# Create a vector of 24 random numbers from a normal distribution
y <- rnorm(24, mean = 0, sd = 1)
```

## A run chart from random normal data

First we plot a simple run chart without any helper lines (Figure \@ref(fig:first-chart-run1)):

```{r first-chart-run1, fig.cap='Simple run chart'}
# Make point-and-line plot
plot(y, type = 'o')
```

(As a side note, this reminds me (JA) of a manager, who once said to me: "You make such beautiful graphs, but can't you stop them from going up and down all the time." &#128513;)

Then we add the the centre line to the plot. Since we know the parameters of the distribution data come from, we can use the theoretical population mean/median (= 0) as the centre line.

With the centre line in place, we can apply runs analysis to identify unnatural patterns suggesting sustained shifts or trends in data over time (Figure \@ref(fig:first-chart-run2)):

```{r first-chart-run2, fig.cap='Run chart with centre line'}
# Create y-coordinates for the centre line
cl  <- rep(0,  length(y))

# Plot data and add centre line
plot(y, type = 'o')
lines(cl)
```

We find that the longest runs (there are two of them, 6:10 and 15:19) have 5 data points and that the curve crosses the centre line 9 times. With 24 data points and using the runs rules proposed previously, the upper limit for longest run is 8 as is (coincidentally) the lower limit for number of crossings. Consequently, these data seem to come from a random process (which we already knew).

## Adding control limits to produce a control chart

We use the same technique to add the lower and upper control limits. 

Remember that the control limits are usually set to $CL \pm 3 SD$, where CL is the centre line and SD is the estimated standard deviation -- that is, the standard deviation of the natural variation in data, not the pooled standard deviation that would include both random and any non-random variation.

Again, since we know the parameters of the distribution, we use the theoretical population standard deviation (= 1) to calculate the control limits.

```{r}
# Create y-coordinates for the control limits
lcl <- rep(-3, length(y))
ucl <- rep(3,  length(y))
```

Finally, we need to expand the y-axis limits to make room not only for the data points but also the lines (Figure \@ref(fig:first-chart-ctrl1)):

```{r first-chart-ctrl1, fig.cap='Standardised control chart'}
# Plot data expanding y-axis to make room for all data and lines
plot(y, type = 'o', ylim = range(y, lcl, ucl))

# Add lines
lines(ucl)
lines(cl)
lines(lcl)
```

As expected, all data points fall within the control limits suggesting only random variation.
 
Now, let's mimic a shift in data by introducing four data points (21:24) from a normal distribution with mean = 3 (Figure \@ref(fig:first-chart-ctrl2)):
 
```{r first-chart-ctrl2, fig.cap='Control chart with data points outside the upper control limit (freaks)'}
# Initialise random number generator to make results reproducible
set.seed(19)

# Introduce a shift in data
y[21:24] <- rnorm(4, mean = 3)

# Plot control chart
plot(y, type = 'o', ylim = range(y, lcl, ucl))
lines(ucl)
lines(cl)
lines(lcl)
```

Note that only one of the four data points from the shift is above the upper control limit. This is a reminder that even large shifts in data are not guaranteed to show up immediately. Rather, shifts in data increases the *chance* that future data point will fall outside the limits. The chance of discovering a shift depends on two things: 1. the size of the shift; 2. how long we are willing to wait for a shift to show itself.

## That's all, Folks!

So constructing an SPC chart using R may be done using only a few lines of code. In fact, most of the code in this chapter went to prepare the data to be plotted. The charts themselves are rather simple and plotting is the same every time: 1. plot the dots; 2. add the lines.

Later we will wrap all the steps in a function that also automates the calculation of empirical centre and control lines, highlights signals of non-random variation in data, and makes plots that are a lot nicer to look at than the rather crude ones we have produces so far.

In the next chapter we will calculate control limits for the control charts that are most commonly used in healthcare, "The Magnificent Seven". This is all about estimating the standard deviation in data, which depends on the type of data.
