---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Core R Functions to Construct SPC Charts {#r-functions}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/3,
                      dev     = 'svg')

source('R/spc_functions.R', local = knitr::knit_global())
```

Until now, we have calculated control limits manually before plotting. In this chapter, we introduce a library of R functions that work together to automate the main steps involved in constructing SPC charts.

We will not go through each of these functions in detail, but we encourage you to study them carefully in order to understand how they work individually and how they work together. We also encourage you to adapt and improve them to suit your own needs.

In total, the library consists of 17 functions. However, the user only needs to interact directly with one of them: `spc()`. The source code for the full function library is provided in the section at the end of this chapter.

The main function, `spc()`, is an extended version of the improved `spc()` function introduced in Chapter \@ref(highlighting).

Most importantly, it is no longer necessary to calculate control limits manually. Instead, we provide a `chart` argument, which should be one of the following: 'run', 'xbar', 's', 'i', 'mr', 'c', 'u', or 'p'. If no chart argument is supplied, the function produces a run chart.

We also no longer need to use the somewhat clumsy $ notation or the `with()` function to access variables inside a data frame. Instead, we can pass the name of the data frame to the data argument.

Finally, for X-bar and S charts, there is no longer any need to aggregate the data in advance. That task is handled by the `spc.aggregate()` function, which is also responsible for calling the appropriate functions to calculate centre lines and control limits and to perform the runs analysis.

After completing these steps, `spc.aggregate()` returns a data frame containing all the information needed to construct the plot. The actual drawing of the chart is then handled by the `plot.spc()` function.

## Examples

This section gives examples of a run chart and each of the *Magnificent Seven*. Documentation for the datasets used in the examples is provided in Appendix \@ref(data-sets).

### Run chart

```{r rfun-run, fig.cap='Run chart'}
d <- read.csv('data/blood_pressure.csv',
              comment.char = '#',
              colClasses   = c(date = 'Date'))

head(d)

spc(date, systolic,
    data = d,
    main = 'Systolic blood pressure',
    ylab = 'mm Hg',
    xlab = 'Date')
```

### I and MR charts for individual measurements

```{r rfun-imr, fig.asp=1, fig.cap='I and MR charts'}
# Setup plotting area to hold two plots on top of each other and adjust margins
op <- par(mfrow = c(2, 1),
          mar   = c(4, 4, 2, 0) + 0.2)
spc(date, systolic, 
    data  = d, 
    chart = 'i',
    main  = 'Systolic blood pressure',
    ylab  = 'mm Hg',
    xlab  = NA)
spc(date, systolic, 
    data  = d, 
    chart = 'mr',
    main  = 'Moving range',
    ylab  = 'mm Hg',
    xlab  = 'Date')
# Reset plotting area to default
par(op)
```

### X-bar and S charts for averages and standard deviations of measurements

```{r rfun-xbars, fig.asp=1, fig.cap='X-bar and S charts'}
d <- read.csv('data/renography_doses.csv',
              comment.char = '#',
              colClasses   = c(date = 'Date',
                               week = 'Date'))

head(d)

op <- par(mfrow = c(2, 1),
          mar   = c(4, 4, 2, 0) + 0.2)
spc(week, dose,
    data  = d,
    chart = 'xbar',
    main  = 'Average radiation dose for renography',
    ylab  = 'MBq',
    xlab  = NA)
spc(week, dose,
    data  = d,
    chart = 's',
    main  = 'Standard deviation',
    ylab  = 'MBq',
    xlab  = 'Week')
par(op)
```

### C and U charts for counts and rates

```{r rfun-cu, fig.asp=1, fig.cap='C and U charts'}
d <- read.csv('data/bacteremia.csv',
         comment.char = '#',
         colClasses   = c(month = 'Date'))

head(d)

op <- par(mfrow = c(2, 1),
          mar   = c(4, 4, 2, 0) + 0.2)
spc(month, ha_infections,
    data = d,
    chart    = 'c',
    main     = 'Hospital-acquired bacteremia',
    ylab     = 'Count',
    xlab     = NA)
spc(month, ha_infections, risk_days,
    data     = d,
    chart    = 'u',
    multiply = 10000,
    main     = NA,
    ylab     = 'Count per 10,000 risk-days',
    xlab     = 'Month')

par(op)
```

### P chart for percentages

```{r rfun-p, fig.cap='P chart'}
spc(month, deaths, patients,
    data     = d,
    multiply = 100,
    chart    = 'p',
    main     = '30-day mortality after bacteremia',
    ylab     = '%',
    xlab     = 'Month')
```

## TODO

One obvious shortcoming of this function library is that it lacks error checking. If you plan to use these functions in a production environment or as part of your own SPC package, you will need to add this yourself. At a minimum, you should check automatically that the inputs are of the expected types and that the x, y, and n arguments have the same lengths. As a starting point, see the `stopifnot()` function.

## Further up, further in

We now have a functioning library of R functions that automates most of the steps involved in constructing SPC charts. There is still plenty of room for improvement, but the library should be enough to get you started and, more importantly, to give you a deeper understanding of the considerations involved in plotting SPC charts.

In the next chapter, we take a brief look at how to use *ggplot2()* for plotting instead of the `plot()` function from base R.

----

## R function library {- #funs}

```{r, file='R/spc_functions.R'}
```

