---
output: html_document
editor_options: 
  chunk_output_type: console
---
# SPC Charts with `ggplot2` {#ggplot}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/5,
                      dev     = 'svg')

source('R/spc_functions.R', local = knitr::knit_global())
```

Armed with the battery of functions from Chapter \@ref(r-functions) we are able to construct any of the Magnificent Eight using only functionality from base R. Furthermore, it is easy to add new types of SPC charts to the library. To achieve this, all we have to do is to write an appropriate `spc.*()` function to handle the calculations of the centre line and limits and to include the function type in the`spc()` function's chart argument.

For pedagogical and practical reasons we chose to use only functions from base R in our library. However, for other reasons one might prefer plotting using a more modern and easily adaptable plotting system like `ggplot2` or `lattice`. Because we have used a modularised approach and made a separate function for plotting spc objects, `plot.spc()`, it is a simple task to use any other plotting method.

First we will show how to create an object of class spc, which we can assign a variable name for later use:

```{r}
d <- read.csv('data/cdiff.csv',
              comment.char = '#',
              colClasses = c(month = 'Date'))

p <- spc(month, infections,
         data  = d,
         chart = 'c',
         plot = FALSE)

head(p)
```

Note that we suppressed the plotting

```{r}
plot(p)
```

```{r}
library(ggplot2)

ggspc <- function(p) {
  col1   <- 'steelblue'
  col2   <- 'tomato'
  dotcol <- ifelse(p$sigma.signal, col2, col1)
  clcol  <- ifelse(p$runs.signal[1], col2, col1)
  cltyp  <- ifelse(p$runs.signal[1], 'dashed', 'solid')
  
  ggplot(p, aes(x, y)) +
    geom_line(aes(y = lcl), na.rm = TRUE) +
    geom_line(aes(y = ucl), na.rm = TRUE) +
    geom_line(aes(y = cl),
              na.rm    = T,
              colour   = clcol,
              linetype = cltyp) +
    geom_line(colour = col1, na.rm = TRUE) +
    geom_point(colour = dotcol, na.rm = TRUE)
}

ggspc(p)
```

```{r}
p <- ggspc(p)

p +
  theme_light() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = 'gray')) +
  labs(title = 'CDiff infections',
       y = 'Count',
       x = 'Month')
```

```{r}
plot.spc <- ggspc

plot(p)
```

