---
output: html_document
editor_options: 
  chunk_output_type: console
---
# SPC Charts with *ggplot2* {#ggplot}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/5,
                      dev     = 'svg')

source('R/spc_functions.R', local = knitr::knit_global())
source('R/load_data.R', local = knitr::knit_global())
```

Armed with the set of functions developed in Chapter \@ref(r-functions) , we are now able to construct all of the most commonly used SPC charts using base R. Furthermore, the modular structure of the function library makes it straightforward to extend. To add a new type of chart, we simply need to write a corresponding spc.*() function to calculate the centre line and control limits, and then include this function in the chart argument of the main `spc()` function.

Because we have separated the calculation of chart components from the plotting itself -- by introducing a dedicated plotting method, `plot.spc()` -- it is also easy to replace the base R graphics engine with an alternative like *ggplot2* or *lattice*.

In this chapter, we develop an alternative plotting function based on *ggplot2*. The advantage of *ggplot2* is not that it can do things that base R cannot, but that it often makes common tasks simpler and more consistent.

For example, *ggplot2* automatically handles scaling of axes and layout of labels, so we do not need to adjust these manually when adding elements to a plot. In addition, *ggplot2* provides a powerful theming system that makes it relatively easy to customise the non-data elements of a chart, such as colours, grid lines, and axis formatting.

## Creating an SPC object for later plotting

Using the `spc()` function from the previous chapter, we can create an SPC object and store it for later use:

```{r}
# make spc object
p <- spc(month, infections,
         data  = cdiff,
         chart = 'c',
         plot  = FALSE)

```

Here, we suppress plotting by setting plot = FALSE. The function instead returns a data frame containing all the coordinates and metadata needed to construct the chart. This object is assigned to p, which we can now manipulate like any other R object.

```{r}
# check the class of p
class(p)

# show the first six rows from p
head(p)
```

Because p is an object of class "spc" we simply call the generic `plot()` function: `plot(p)`.

```{r ggplot-fig1, fig.cap='SPC chart'}
# show the C chart
plot(p) # not necessary to call spc.plot(), just call the generic plot() function
```

The generic `plot()` function automatically dispatches to the specialised `plot.spc()` method (\@ref(fig:ggplot-fig1)).

## Making a new plot function based on *ggplot2*

We can now define an alternative plotting function that uses *ggplot2*.

```{r}
# Load ggplot2
library(ggplot2)

# Function for plotting spc objects with ggplot()
ggspc <- function(p) {
  # Set colours
  col1    <- 'steelblue'
  col2    <- 'tomato'
  linecol <- 'gray50'
  dotcol  <- ifelse(p$sigma.signal, col2, col1)
  clcol   <- ifelse(p$runs.signal[1], col2, linecol)
  cltyp   <- ifelse(p$runs.signal[1], 'dashed', 'solid')
  
  # Plot the dots and draw the lines
  ggplot(p, aes(x, y)) +
    geom_line(aes(y = lcl), colour = linecol, na.rm = TRUE) +
    geom_line(aes(y = ucl), colour = linecol, na.rm = TRUE) +
    geom_line(aes(y = cl), colour = clcol, linetype = cltyp, na.rm = TRUE) +
    geom_line(colour = col1, na.rm = TRUE) +
    geom_point(colour = dotcol, na.rm = TRUE)
}
```

We then use this function to plot the SPC object (Figure \@ref(fig:ggplot-fig2)).

```{r ggplot-fig2, fig.cap='SPC chart using the `ggspc()` function'}
# Plot an spc object
ggspc(p)
```

This function produces a plot equivalent to the base R version, but using *ggplot2*’s grammar of graphics.

We can also convert the SPC object into a ggplot object and then further customise it as in Figure \@ref(fig:ggplot-fig3).

```{r ggplot-fig3, fig.cap='SPC chart with modified theme'}
# convert to ggplot2 object
p <- ggspc(p)
class(p)

# modify theme and labels
p +
  theme_light() +
  theme(panel.grid   = element_blank(),
        panel.border = element_blank(),
        axis.line    = element_line(colour = 'gray')) +
  labs(title = 'CDiff infections',
       y     = 'Count',
       x     = 'Month')
```

At this point, we could replace the existing `plot.spc()` function with this new implementation if desired. The choice between base R and *ggplot2* ultimately comes down to personal preference and use case.

## Customising the plotting theme

The final example in this chapter shows how to define a custom theme and format the y-axis as percentages (Figure \@ref(fig:ggplot-fig4)).

```{r ggplot-fig4, fig.cap='SPC chart with labels and custom y-axis'}
mytheme <- function() {
  theme_light() +
  theme(panel.grid   = element_blank(),
        panel.border = element_blank(),
        axis.line    = element_line(colour = 'gray'))
}

p <- spc(month, deaths, patients,
          data = bact,
          chart = 'p',
          plot = FALSE)

ggspc(p) +
  mytheme() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = '30-day mortality',
       y     = NULL,
       x     = 'Month')

```

In practice, repeatedly defining themes and formatting axes for each plot can become tedious. Ideally, these elements would be handled automatically.

## Preparing for *qicharts2*

This is precisely the aim of *qicharts2*, an R package for SPC charts that we introduce in the next chapter. The package builds on the same principles developed so far but provides additional functionality for customisation and automation.

In particular, qicharts2 makes it easy to create multidimensional plots using *ggplot2*’s faceting capabilities, allowing multiple related charts to be displayed together in a consistent and flexible way.
