---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Introducing qicharts2 {#qicharts}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/5,
                      dev     = 'svg')
source('R/load_data.R', local = knitr::knit_global())
library(qicharts2)
```

qicharts2 (Quality Improvement Charts, @qicharts2) is an R package for SPC aimed at healthcare data analysts. It is based on the same principles that we have developed in the previous chapters of this book. It contains functions to construct all of the Magnicent Eight plus a number of specialised charts including pareto charts and control charts for rare events data.

To learn everything about qichart2, visit its website: https://anhoej.github.io/qicharts2/. In this chapter we will concentrate on some key facilities that is still missing from the function library we have buld so far:

* excluding data points from analysis
* freezing and splitting charts by periods
* multivariate plots (small multiples)

To get started with qicharts2, install it and load it into your working environment:


```{r, eval=FALSE}
install.packages('qicharts2')
library(qicharts2)
```

As always, you only need to install a package once, but you need to load it every time you want to use it. 

Next, you may want to read the vignette: `vignette('qicharts2')` -- or you may want to get started right away.

## A simple run chart

The main function of qicharts2 is `qic()`. It takes the same arguments as the `spc()` function we built previously plus many more. Check the documentation for a complete list of arguments: `?qic`.

To reproduce our first run chart from Chapter \@ref(first-chart), run:

```{r}
qic(systolic)
```

There are several things to note:

* Default chart title and axis labels are created automatically. These may, of course, be changed using the `title`, `ylab`, and `xlab` arguments.

* Data points that fall directly on the centre line are greyed out. These do not count as useful observation in the runs analysis.

* The centre line value is printed on the chart.

## A simple control chart

To produce a control chart, we simply add a `chart` argument:

```{r}
qic(systolic, chart = 'i')
```

By default `qic()` uses a grey background area to show the natural process limits.

## Excluding data points from analysis

Sometimes it is useful to exclude one or more data points from the calculation of control limits and from the runs analysis. This may be the case when specific data points are known to have been influenced by factors that are not part of the natural process, for example data points that fall outside the control limits. We use the `exclude` argument to do this:

```{r}
qic(systolic, chart = 'i', exclude = 6)
```

Notice how the values for the centre line and control limits changed a little. Specifically, the control limits became slightly narrower.

Excluding data points should be a deliberate decision and not something that is done automatically just because one or more data points are outside the control limits. Exclusion should be based on a thorough understanding of the process and only when the reason(s) for a special cause has been established. Otherwise, the whole idea of SPC as a way of understanding variation and its sources is lost.

## Freezing baseline period

When data have been collected for a long enough period of time to establish the centre line and control limits of a stable and predictable process -- that is, a process with only common cause variation -- it is often useful to "freeze" the limits and use them for future data.

In production industry this technique is referred to as phase one and phase two studies. In healthcare freezing is especially useful when we have historical data from before the start of some type of intervention or improvement programme. When plotting future data with the centre line and control limits from a stable baseline period shifts and trends will show up faster than if they were recalculated with every new data point.

The cdi dataset comes with qicharts2 and contains monthly counts of infections two years before and one year after the initiation of an improvement programme. Check the documentation for details, `?cdi`.

```{r qicharts-cdi1, fig.cap='Infections before and after intervention'}
qic(month, n, data = cdi, freeze = 24)
```

Figure \@ref(fig:qicharts-cdi1) shows a run chart of data where the centre line has been established from the baseline period (month 1-24) using the freeze argument. There is a sustained shift in data after the intervention.

## Splitting chart by period

When a sustained shift in data has been discovered *and* the cause is known it is allowable to split the graph by periods. We use the part argument for this (Figure \@ref(fig:qicharts-cdi2)).

```{r qicharts-cdi2, fig.cap='Splitting using index'}
qic(month, n, data = cdi, part = 24)
```

The split argument takes either the indices of the data points to split after, or we may use a categorical variable naming the time periods in which case the periods will be labelled (Figure \@ref(fig:qicharts-cdi3)).

```{r qicharts-cdi3, fig.cap='Splitting using a period variable'}
qic(month, n, data = cdi, part = period)
```

As with excluding data points, freezing and splitting should be based on deliberate decisions and thorough understanding of the process. Some SPC applications allow for automatic splitting whenever a shift is detected. We strongly advise against this approach. Freezing and splitting *may* be useful when:

* there is a sustained shift in data
* the reason for the shift is known
* the shift is in the desired direction
* the shift is expected to continue

If any of these conditions is not met we should rather look for the root cause and -- if need be -- eliminate it.

## Small multiple plots for multivariate data

Process data is all about time. But often data -- not the least in healthcare -- have more dimensions, which are important to understand in order to interpret data correctly.

- one unit, many indicators

- one indicator, many units

- many units, many indicators






For example, the hospital_infections dataset, which is also included in the qicharts2 package, contains monthly counts of three types of hospital infections, bacteremia (BAC), C. diff (CDI), and urinary tract infections (UTI) from six hospitals.

```{r}
# show structure of hospitals_infections dataset
str(hospital_infections)
```

In this example data have two extra dimensions, infections and hospitals.

A most useful way of visualising multivariate data is to plot data from different "spaces" alongside each other. 

```{r qicharts-hais1, fig.cap='Small multiple plot of urinary tract infection from six hospital'}
qic(month, n, days, 
    data = subset(hospital_infections, infection == 'UTI'),
    chart = 'u',
    multiply = 10000)
```

```{r, fig.asp=3/3}
qic(month, n, days, 
    data = subset(hospital_infections, infection == 'UTI'),
    chart = 'u',
    multiply = 10000,
    facets = ~hospital)
```

```{r}
qic(month, n, days, 
    data = subset(hospital_infections, hospital == 'NOH'),
    chart = 'u',
    multiply = 10000,
    facets = ~infection)
```

```{r, fig.asp=3/3}
qic(month, n, days, 
    data = subset(hospital_infections, hospital == 'NOH'),
    chart = 'u',
    multiply = 10000,
    facets = ~infection,
    scales = 'free_y',
    ncol = 1)
```

```{r}
qic(month, n, days, 
    data = hospital_infections,
    chart = 'u',
    multiply = 10000,
    facets = infection ~ hospital,
    scales = 'free_y',
    title = 'Hospital infections',
    ylab = 'Count per 10,000 risk days',
    xlab = 'Month')
```
