# (PART\*) Part 3: Advanced SPC Techniques {-}

# Screened I Chart -- eliminating freak moving ranges before calculating limits {#screenedmr}

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/4,
                      dev     = 'svg')

library(qicharts2)

set.seed(44)
y   <- rpois(24, 16)
mr <- c(NA, abs(diff(y)))
amr <- mean(mr, na.rm = TRUE)
mr_screened <- mr
mr_screened[mr >= 3.267 * amr] <- NA
amr_screened <- mean(mr_screened, na.rm = T)

m <- rbind(y, mr)
colnames(m) <- seq_along(y)
m <- cbind(m, avg = round(rowMeans(m, na.rm = TRUE), 2))
```

As discussed in earlier chapters, control limits are designed to represent natural (common cause) variation and thereby enclose almost all data points from a stable process. Estimation of this variation typically relies on within-subgroup variation. However, when subgroups consist of single observations -- as in I charts -- there is no within-subgroup variation to calculate. Instead, we use moving ranges, defined as the absolute differences between consecutive data points. In effect, pairs of consecutive observations are used to estimate the within-subgroup variation.

This approach has an important implication. If a shift occurs between two consecutive observations, the moving range at that point will be inflated and may produce a signal on the moving range chart.

To illustrate, consider the following 24 values:

```{r, echo=FALSE, comment=''}
cat(y)
```

The average moving range (AMR) is `r amr`. From Table \@ref(tab:limits-tab1), the natural upper limit for moving ranges is 3.267 x AMR (= `r round(3.267 * amr, 1)`). The moving range between observations 14 and 15 is |28 − 11| = 17, which exceeds this limit and is therefore unusually large, as shown in Figure \@ref(fig:screened-mr1).

```{r screened-mr1, fig.cap='Moving range chart with one range above the control limit'}

qic(y, 
    chart = 'mr', 
    title = 'Moving range chart')
```

Some authors recommend removing such extreme moving ranges before calculating the control limits for the corresponding I chart, as this may improve sensitivity to meaningful changes [@nelson1982]. By default, qicharts2 applies this approach, as illustrated in Figure \@ref(fig:screened-mr2).

```{r screened-mr2, fig.cap='I chart with control limits calculated after removing freak moving ranges'}
# default is to screen moving ranges before calculating control limits
qic(y, 
    chart = 'i', 
    title = 'I chart with screened MRs')
```

The procedure is as follows. First, calculate the AMR from the original data (= `r amr`). Next, remove any moving ranges greater than 3.267 × AMR (= `r 3.267 * amr`) and recalculate the AMR (= `r round(amr_screened, 2)`). Finally, use this screened AMR to calculate the control limits for the I chart: (= 15.8 $\pm$ 2.66 x 4.45).
 
To suppress this screening step in qicharts2, the qic.screenedmr option can be set to FALSE as in Figure \@ref(fig:screened-mr3):

```{r screened-mr3, fig.cap='I chart with control limits calculated without removing extreme ranges'}
# suppress screening of moving ranges
options(qic.screenedmr = FALSE)
qic(y, 
    chart = 'i',
    title = 'I chart without screened MRs')

# reset screening option to default
options(qic.screenedmr = TRUE)
```

As expected, the control limits in the unscreened chart are slightly wider than those in the screened chart -- in this example, just wide enough to suppress a signal that would otherwise have been detected.

One might argue that screening moving ranges is unnecessary if the moving range chart is presented alongside the I chart. However, in our experience, moving range charts rarely add value for non-technical users and may even introduce confusion. For this reason, we prefer to increase the sensitivity of I charts by excluding extreme moving ranges before calculating control limits, while reserving moving range charts for more technical analysis.
