# (PART\*) Part 4: Advanced SPC Techniques {-}

# Screened I Chart -- eliminating freak moving ranges before calculating limits

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/4,
                      dev     = 'svg')

library(qicharts2)

set.seed(44)
y   <- rpois(24, 16)
amr <- mean(abs(diff(y)))
```

As detailed in earlier chapters, control limits are designed to represent the natural, common cause variation and thereby enclose (almost) all data points from a stable process. To estimate the common cause variation we use the within subgroup variation. However, if the subgroups consist of only one data element, as is the case with I charts, there is no within subgroup variation to be calculated. Instead, we use the moving ranges, that is the absolute differences between consecutive data points. Thus, *two* subgroups is used to define the "within" subgroup variation.

Consequently, if a shift occurs between two subgroups, this will affect the moving range at this point and possible produce a signal on the moving range chart. For example, take these 24 "random" numbers:

```{r, echo=FALSE, comment=''}
cat(y)
```

The average moving range (AMR) is `r amr`, and the upper control AMR limit is 3.267 x `r amr` = `r round(3.267 * amr, 1)`.

The range between observations numbers 14 and 15 (|11 -- 28| = 17) is unusually large as shown by the moving range chart in Figure \@ref(fig:screened-mr1).

```{r screened-mr1, fig.cap='Moving range chart with one range above the control limit'}

qic(y, 
    chart = 'mr', 
    title = 'Moving range chart')
```

Some authors recommend removing extreme moving ranges before calculating the control limits for the corresponding I chart [@nelson1982], as this can slightly improve the chart's sensitivity. qicharts2 does this by default as shown in Figure \@ref(fig:screened-mr2).

```{r screened-mr2, fig.cap='I chart with control limits calculated after removing freak moving ranges'}
# default is to screen moving ranges before calculating control limits
qic(y, 
    chart = 'i', 
    title = 'I chart with screened MRs')
```

One may argue that this is unnecessary as long as the moving range chart is shown alongside the I chart. To suppress the screening of moving ranges before calculating control limits for I charts (and prime charts) in qicharts2, we can set the qic.screenedmr option to FALSE before plotting as in Figure \@ref(fig:screened-mr3):

```{r screened-mr3, fig.cap='I chart with control limits calculated without removing ranges above control limit'}
# suppress screening of moving ranges
options(qic.screenedmr = FALSE)
qic(y, 
    chart = 'i',
    title = 'I chart without screened MRs')

# reset screening option to default
options(qic.screenedmr = TRUE)
```

In our experience, presenting moving range charts to non-technical SPC users rarely adds meaningful insight and can sometimes cause confusion. As a result, we prefer to enhance the sensitivity of I charts by excluding extreme moving ranges before calculating control limits while leaving the use and interpretation of moving range charts to data analyst.
