# Prime Charts for Count Data with Very Large Subgroups

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/4,
                      dev     = 'svg')

library(qicharts2)
```

With count data involving very large subgroup sizes, SPC charts often produce very tight control limits with many data points outside the limits. 
Figure \@ref(fig:pcharts-fig1) demonstrates this phenomenon, known as overdispersion, using data from from @mohammed2013 included in qicharts2 (data are tabulated at the [end of this chapter](#pp-data)).

```{r pcharts-fig1, echo=FALSE, fig.cap='P chart of percent emergency attendances seen within 4 hours'}
qic(i, r, n, 
    data     = nhs_accidents,
    chart    = 'p',
    title    = 'Attendances seen within 4 hours',
    subtitle = 'P chart',
    xlab     = 'Week')
```

This can suggest that the process is highly unstable, even when such a conclusion may not be warranted. Overdispersion occurs when the natural, common cause variation in the data exceeds what is expected under the assumed theoretical distribution -- binomial for proportions and Poisson for rates.

To diagnose this issue, we can plot the data on an I chart, which bases its control limits on the natural variation between successive subgroups rather than on theoretical distributional assumptions. Figure \@ref(fig:pcharts-fig2) is an I chart of the data from Figure \@ref(fig:pcharts-fig1) suggesting only common cause variation.

```{r pcharts-fig2, echo=FALSE, fig.cap='I chart of percent emergency attendances seen within 4 hours'}
qic(i, r, n, 
    data      = nhs_accidents,
    chart     = 'i',
    y.percent = TRUE,
    title     = 'Attendances seen within 4 hours',
    subtitle  = 'I chart',
    xlab      = 'Week')
```

As a general rule of thumb, when the control limits from an I chart differ significantly from those of the P or U chart, this suggests that the observed variation is inconsistent with the assumed binomial or Poisson models underlying the P and U charts, respectively.

Several solution to the problem with overdispersion and tight control limits have been proposed. The obvious pragmatic solution is to simply use I charts for count data as well as for (individual) measurements. However, I charts have straight control limits that do not account for varying subgroup sizes. If the subgroup sizes only vary little, this may not be a problem. But sometimes -- not the least in healthcare -- varying subgroup sizes matter.

Laney proposed a solution that adjusts the assumed within subgroup variation in count data by a factor based on the moving ranges of successive standardised data points ($\sigma_z$) [@laney2002].

## Laney's prime chart

As shown earlier in Table \@ref(tab:limits-tab1), the control limits for a **traditional P chart** are calculated as:

$$\bar{p}\pm3\sigma_{pi}$$

where $\bar{p}$ is the average proportion and $\sigma_{pi}$ is the within-subgroup standard deviation for the i-th subgroup:

$$\sigma_{pi}=\sqrt{\bar{p}(1-\bar{p})/{n_{i}}}$$

where $n_i$ is the sample size of the i-th subgroup.

The **P′ chart** (pronounced P prime chart) accounts for both within and between subgroup variation. The control limits are given by:

$$\bar{p}\pm3\sigma_{pi}\sigma_z$$

where $\sigma_z$ represents the between-subgroup variation, calculated using the moving ranges of the standardised proportions defined as:

$$z_i=(p_i - \bar{p})/\sigma_{pi}$$

The moving ranges are:

$$MR_i=|z_i-z_{i-1}|$$

Then, the estimated between-subgroup variation $\sigma_z$ is the average moving range devided by the constant 1.128:

$$\sigma_z=\overline{MR_z}/1.128$$
Thus, the control limits for the P′ chart are:

$$\bar{p}\pm3\sigma_{pi}\sigma_z $$

```{r pcharts-fig3, echo=FALSE, fig.cap="P' chart of percent emergency attendances seen within 4 hours"}
qic(i, r, n, 
    data      = nhs_accidents,
    chart     = 'pp',
    y.percent = TRUE,
    title     = 'Attendances seen within 4 hours',
    subtitle  = "P' chart",
    xlab      = 'Week')
```

The control limits in Figure \@ref(fig:pcharts-fig3) are very close to the I chart limits from Figure \@ref(fig:pcharts-fig3), but vary slightly due to varying subgroup sizes.

Similarly, control limits for the U' chart are:

$$\bar{u}\pm3\sigma_{ui}\sigma_z$$

where $\sigma_{ui}=\sqrt{\bar{u}/n_i}$ and $\sigma_{z}=\overline{MR_z}/1.128$; and the moving ranges are computed from the standardised rates defined as $z_i=(u_i-\bar{u})/\sigma_{ui}$.

Note that in qicharts2, the moving ranges of the standardized values are, by default, screened for extreme values -- following the same approach used for I charts, as described in Chapter \@ref(screenedmr).

## When to use prime charts

Laney's prime charts were developed to handle overdispersion (and underdispersion), that is, when data show more (or less) variation than expected under a binomial or Poisson distribution, which is often seen with very large subgroup sizes. But how do we know when to suspect overdispersion?

A red flag is when a traditional P or U chart shows unusually tight control limits that seem unnatural. If plotting the same data on an I chart produces significantly wider control limits, it is a strong indication that the assumptions behind the P or U chart may not hold. In such cases, switching to Laney’s P' or U' charts is typically the better choice.

Note that when the adjustment factor ($\sigma_z$) is close to one -- indicating little to no overdispersion -- the Laney P′ or U′ chart closely resembles the traditional P or U chart. For this reason, Laney recommends using prime charts as a generally safe and robust default.

In the next chapter, we will explore a modified I chart -- the I' (I prime) chart -- which generates control limits for count data that closely align with Laney’s prime charts, while also accommodating measurement data with varying subgroup sizes.

----

## Example data for P prime charts {- #pp-data}

```{r pcharts-tab1, echo=FALSE}
d <- nhs_accidents
names(d) <- c('Week (i)',
              'Attendances seen within 4 hours (r)',
              'Number of attendances (n)')
knitr::kable(d, 
             format.args = list(big.mark = ','),
             caption = 'The number of attendances to major accident and emergency hospital departments in the NHS that were seen within 4 hours of arrival over twenty weeks. Source: [@mohammed2013]')
```

