---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Funnel Plots for Categorical Subgroups

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/4,
                      dev     = 'svg')

library(qicharts2)
```

Funnel plots are SPC charts with categorical subgroups, which are often sorted after size to produce funnel shaped control limits. Because there is no natural time order, funnel plots have no connecting lines between data points and runs analysis is also not useful.

Funnel plots are useful for comparing indicator levels from different categories, for example infection rates from different hospitals.

We use data on urinary tract infections from the `hospital_infections` dataset from qicharts2.

```{r funnel-fig1, fig.asp=1, fig.cap='U chart of urinary tract infections'}
# get urinary tract infections from hospital infections data
uti <- hospital_infections[hospital_infections$infection == 'UTI',]
# plot U chart
qic(month, n, days,
    data     = uti,
    chart    = 'u',
    multiply = 10000,
    facets   = ~hospital,
    ncol     = 2,
    title = 'Hospital acquired urinary tract infections in six hospitals',
    ylab = 'Counts per 10,000 risk days',
    xlab = 'Month')
```

Figure \@ref(fig:funnel-fig1) is a traditional U chart of infection rates in six hospital. There are signals of special cause variation in two hospitals (BOH, NOH), which makes direct comparison of infection rates between hospitals pointless -- it makes no sense to compare levels from data that are moving.

It is therefore important to select data from a recent and stable period of time before constructing funnel plots. In Figure \@ref(fig:funnel-fig2) we have selected data from the most recent quarter and plotted them using hospital as the subgrouping variable.

```{r funnel-fig2, fig.cap="Unordered 'funnel' plot of urinary tract infections"}
# filter data to latest quarter
uti_2016q4 <- uti[uti$month >= '2016-10-01', ]
# plot "funnel" plot
qic(hospital, n, days,
    data     = uti_2016q4,
    chart    = 'u',
    multiply = 10000,
    title = 'Hospital acquired urinary tract infections in six hospitals',
    ylab = 'Counts per 10,000 risk days',
    xlab = 'Hospital')
```

By default the x-axis will be sorted alphabetically. To produce a funnel plot, we need to sort the subgroups by size as is shown in Figure \@ref(fig-funnel-fig).

```{r funnel-fig3, fig.cap='Ordered funnel plot of urinary tract infections'}
# order hospitals by denominator (risk days) to produce funnel
qic(reorder(hospital, days), n, days,
    data     = uti_2016q4,
    chart    = 'u',
    multiply = 10000,
    title = 'Hospital acquired urinary tract infections in six hospitals',
    ylab = 'Counts per 10,000 risk days',
    xlab = 'Hospital')
```


