---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Funnel Plots for Categorical Subgroups

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/5,
                      dev     = 'svg')

library(qicharts2)
```

Funnel plots are SPC charts used for categorical subgroups [@spiegelhalter2005]. The subgroups are often ordered by size, producing the characteristic funnel-shaped control limits. Funnel plots are useful for comparing indicator levels across categories, for example complication rates between hospitals.

Figure \@ref(fig:funnel-fig0) compares the level of an indicator across six organisational units. One unit (E) falls below the lower control limit, suggesting that its results may reflect a fundamentally different process compared with the others.

```{r funnel-fig0, fig.cap='Funnel plot', echo=FALSE}
f <- hospital_infections[hospital_infections$infection == 'UTI' &
                           hospital_infections$month >= '2016-10-01',]

f$hospital <- factor(f$hospital, labels = LETTERS[1:6])

qic(reorder(hospital, days), n, days,
    data     = f,
    chart    = 'u',
    multiply = 10000,
    title = NULL,
    ylab = 'Indicator',
    xlab = 'Subgroup')
```

Because funnel plots do not involve a natural time sequence, the data points are not connected, and run analysis is not applicable.

To demonstrate the construction of funnel plots, we use data on urinary tract infections from the hospital_infections dataset included in qicharts2. We begin by plotting the data over time using a traditional U chart:

```{r funnel-fig1, fig.asp=1, fig.cap='U chart of urinary tract infections'}
# get urinary tract infections from hospital infections data
uti <- hospital_infections[hospital_infections$infection == 'UTI',]

# plot U chart
qic(month, n, days,
    data     = uti,
    chart    = 'u',
    multiply = 10000,
    facets   = ~hospital,
    ncol     = 2,
    title    = 'Hospital acquired urinary tract infections in six hospitals',
    ylab     = 'Infections per 10,000 risk days',
    xlab     = 'Month')
```

Figure \@ref(fig:funnel-fig1) shows signals of special cause variation in two hospitals (BOH and NOH). In such cases, direct comparison of infection rates between hospitals is problematic, as it does not make sense to compare levels based on data that are changing over time.

For this reason, we select data from a recent period before constructing the funnel plot. In Figure \@ref(fig:funnel-fig2), the data are restricted to the most recent quarter and plotted using hospital, rather than time, as the subgrouping variable:

```{r funnel-fig2, fig.cap="Unordered 'funnel' plot of urinary tract infections"}
# filter data to latest quarter
uti_2016q4 <- uti[uti$month >= '2016-10-01', ]

# plot "funnel" plot
qic(hospital, n, days,
    data     = uti_2016q4,
    chart    = 'u',
    multiply = 10000,
    title    = 'Hospital acquired urinary tract infections in six hospitals',
    ylab     = 'Infections per 10,000 risk days',
    xlab     = 'Hospital')
```

By default, the subgroups are ordered alphabetically along the x-axis. To produce a funnel plot, the subgroups must be ordered by size, as shown in Figure \@ref(fig:funnel-fig3).

```{r funnel-fig3, fig.cap='Ordered funnel plot of urinary tract infections'}
# order hospitals by denominator (risk days) to produce funnel
qic(reorder(hospital, days), n, days,
    data     = uti_2016q4,
    chart    = 'u',
    multiply = 10000,
    title    = 'Hospital acquired urinary tract infections in six hospitals',
    ylab     = 'Infections per 10,000 risk days',
    xlab     = 'Hospital')
```

In summary, funnel plots are SPC charts for categorical subgroups and are useful for comparing indicator levels across organisational units. To support meaningful comparison, it is important to use data from a recent period before constructing the funnel plot.
