---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Pareto Charts for Ranking Problems

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/4,
                      dev     = 'svg')

library(qicharts2)

source('R/load_data.R', local = knitr::knit_global())

ae.tbl <- as.data.frame(table(ae$category))
names(ae.tbl) <- c('category', 'count')
```

The Pareto chart, named after [Vilfred Pareto](https://en.wikipedia.org/wiki/Vilfredo_Pareto), was invented by [Joseph M. Juran](https://en.wikipedia.org/wiki/Joseph_M._Juran) as a practical tool for identifying the most important causes of a problem. It is widely used in quality improvement to prioritise efforts by highlighting the categories that contribute most to an outcome.

In this example, we use the dataset on [adverse events](#adverse-events) causing harm to patients, collected using the Global Trigger Tool method [@plessen2012].

```{r}
# print structure of ae data
str(ae)
```

The `paretochart()` function from the qicharts2 package takes a categorical vector as input and produces a Pareto chart as shown in Figure \@ref(fig:pareto-fig1).

```{r pareto-fig1, fig.cap="Pareto chart of patient harm."}
paretochart(ae$category)
```

In a Pareto chart, the bars represent the counts in each category, while the curve shows the cumulative percentage across categories. In this example, almost 80% of harms arise from just three categories: gastrointestinal, infection, and procedure.

Figure \@ref(fig:pareto-fig2) shows a Pareto chart of harm severity. This chart illustrates that nearly all events resulted in temporary harm (categories E–F).

```{r pareto-fig2, fig.cap="Pareto chart of harm severity: E-I, where E-F = temporary harm, G-H = permanent harm, and I = fatal harm."}
paretochart(ae$severity)
```

The `paretochart()` function expects a character or factor vector as input. However, data are often already aggregated into tabular form:

```{r, comment=''}
ae.tbl
```

To construct a Pareto chart from tabulated data, we must first convert the counts back into a vector. This can be done using the `rep()` function, which repeats each category according to its frequency:

```{r pareto-fig3, fig.cap="Pareto chart constructed from tabular data."}
# make vector from counts
ae.cat <- rep(ae.tbl$category, ae.tbl$count)

# show first six rows of vector
head(ae.cat)

# plot Pareto chart
paretochart(ae.cat)
```

In summary, the Pareto chart is a simple but powerful tool for identifying the most common causes of a problem. In many situations, a large proportion of problems can be attributed to a relatively small number of causes. In the example above, reducing gastrointestinal harm (most often obstipation) and hospital-acquired infections would substantially reduce the overall rate of adverse events.

