---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Pareto Charts for Ranking Problems

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 3/4,
                      dev     = 'svg')

library(qicharts2)

source('R/load_data.R', local = knitr::knit_global())

ae.tbl <- as.data.frame(table(ae$category))
names(ae.tbl) <- c('category', 'count')
```

The Pareto chart, named after [Vilfred Pareto](https://en.wikipedia.org/wiki/Vilfredo_Pareto), was invented by [Joseph M. Juran](https://en.wikipedia.org/wiki/Joseph_M._Juran) as a tool to identify the most important causes of a problem.

For this example, we use the dataset on [adverse events](#adverse-events) causing harm to patients, collected during a number of Global Trigger Tool reviews [@plessen2012].

```{r}
# print structure of ae data
str(ae)
```

The paretochart() function (from qicharts2) takes a categorical vector as argument and plots a Pareto chart as demonstrated in Figure \@ref(fig:pareto-fig1).

```{r pareto-fig1, fig.cap="Pareto chart of patient harm."}
paretochart(ae$category)
```

The bars show the number in each category, and the curve shows the cumulated percentage over categories. Almost 80% of harms come from 3 categories: gastrointestinal, infection, and procedure.

Figure \@ref(fig:pareto-fig2) is a Pareto chart of harm severity demonstrating that nearly all events resulted in temporary harm (E-F).

```{r pareto-fig2, fig.cap="Pareto chart of harm severity: E-I, where E-F = temporary harm, G-H = permanent harm, and I = fatal harm."}
paretochart(ae$severity)
```



```{r, comment=''}
ae.tbl
```

```{r}
ae.cat <- rep(ae.tbl$category, ae.tbl$count)
head(ae.cat)
```

```{r pareto-fig3, fig.cap=""}
paretochart(ae.cat)
```

