---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Common Pitfalls to Avoid

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.asp = 2/4,
                      dev     = 'svg')

library(qicharts2)
```

While SPC is a powerful framework for data-driven quality improvement, its effectiveness can be undermined by a range of common missteps. Misinterpretation of data, misuse of run and control charts, and faulty assumptions can lead to wasted effort or missed opportunities, and may ultimately erode trust in SPC itself.

This chapter outlines several common pitfalls to avoid.

## Data issues

SPC charts are only as good as the data behind them. Poor data quality, including inaccurate measurements, inconsistent definitions or data collected under conditions that vary over time or between locations can significantly compromise their reliability.

How the subgroups are formed is especially important. Subgroups consist of the data elements included in each data point. The way data are sampled and grouped can greatly affect their ability to distinguish common cause from special cause variation. If subgrouping is not done properly, the SPC chart may generate false signals or obscure important ones. We have devoted a full chapter to [rational subgrouping](#subgrouping).

Time spend developing operational indicators definitions and rational subgroups often yields significant returns. Investing effort upfront ensures that data are meaningful and comparable, enabling better decision-making.

## Signal fatigue from over-sensitive SPC rules

Originally, control charts employed only one rule, the 3-sigma rule. However, to increase the chart's sensitivity to minor sustained shifts and other patterns of non-random variation, a plethora of supplementary rules have since been developed. To name a few, we have the Western Electric rules (four rules), the Nelson rules (ten rules), and the Westgard rules (six rules). All are based on identifying non-random patterns in the position of data points relative to the one-, two-, or three-sigma limits, each with different diagnostic properties.

Furthermore, several distinct sets of rules for runs analysis with run charts have also been published. These are based on identifying non-random patterns relative the the centre line.

As discussed earlier in this book, it may seem tempting to apply as many rules as possible to increase the chance of detecting any signs of special cause variation. However, this strategy comes at a price. While additional rules do increase the sensitivity of SPC analysis, they also raise the risk of false alarms. This topic is further explored in Appendix \@ref(diagnostics), [Two Types of Errors When Using SPC].

Apart from the risk of applying too many rules, the rules themselves may also be either overly sensitive or insufficiently sensitive. A prominent example is a commonly used set of runs rules -- including four tests for trends, shifts, and too many or too few runs respectively -- adopted by many healthcare organisations [@provost2011; @perla2011]. This particular set has been shown to produce a very high rate of false signals. For instance, in a run chart with 24 data points, the expected false positive rate is close to 50% [@anhoej2015]. The main contributor appears to be the trend rule -- defined as five or more consecutive data points all increasing or decreasing. Short trends like this are very common, even in sequences of purely random numbers. Moreover, trend rules in general have been thoroughly studied and found to be, at best, unhelpful and, at worst, misleading [@davis1988].

To make matters worse, some practitioners deliberately tighten control limits in an attempt to increase the sensitivity of the sigma rule. This behaviour often stems from another common mistake: using the overall standard deviation to calculate control limits [@wheeler2010; @wheeler2016]. As emphasised repeatedly throughout this book, the overall variation is not appropriate for this purpose, as it includes both common cause and special cause variation -- ultimately inflating the control limits. Inflated limits can, in turn, prompt misguided attempts to narrow them artificially. Instead, control limits, set at ±3-sigma from the centre line, should be based on the estimated within-subgroup standard deviation from rationally formed subgroups. This approach captures common cause variation and provides a sound basis for meaningful process monitoring and improvement.

To avoid signal fatigue -- which, at best, results in wasted effort from chasing false signals -- we recommend that practitioners select their rules carefully, focusing on capturing patterns that are most likely to signal special cause variation in their specific context. The three rules suggested in this book -- the two runs rules and the 3-sigma rule -- have been deliberately designed to detect patterns indicative of one or more of the three types of variation most commonly encountered in healthcare improvement, shifts, trends, or cycles, while keeping acceptable false alarm rates.

## Automatic rephasing

Rephasing involves splitting an SPC chart to recalculate the centre line and control limits after a sustained shift has been identified.

Figure \@ref(fig:pitfalls-fig1) shows the number of hospital infections (C. diff.) before and after an intervention, which began after month 24.

```{r pitfalls-fig1, fig.cap='Hospital infections.'}
qic(month, n,
    data  = cdi,
    chart = 'c',
    title = 'Hospital associated C. diff.-infections',
    ylab  = 'Count',
    xlab  = 'Month')
```

Figure \@ref(fig:pitfalls-fig2) has been rephased following the intervention and the subsequent shift. This correctly distinguishes two distinct periods, each with stable but differing levels of infections.

```{r pitfalls-fig2, fig.cap='Hospital infections -- rephasing done right!'}
qic(month, n,
    data  = cdi,
    chart = 'c',
    part  = period,
    title = 'Hospital associated C. diff.-infections',
    ylab  = 'Count',
    xlab  = 'Month')
```

Automatic rephasing is when the centre line and control limits are recalculated automatically after the software detects a shift -- typically triggered by a prolonged run of data points on one side of the centre line. Some software packages perform automatic rephasing by default (@reading2021).

Figure \@ref(fig:pitfalls-fig3) has been automatically rephased following detection of the shift. Coincidentally, the run that signalled the shift began three months earlier. However, these months occurred prior to the intervention and should correctly be considered part of the baseline period. As a result, the control limits and centre lines do not accurately reflect the distinct before-and-after periods. The centre lines are inflated, and a freak data point falsely suggests instability during the baseline period.

```{r pitfalls-fig3, fig.cap='Hospital infections -- rephasing done wrong!'}
qic(month, n,
    data        = cdi,
    chart       = 'c',
    part        = 22,
    part.labels = c('pre', 'post'),
    title       = 'Hospital associated C. diff.-infections',
    ylab        = 'Count',
    xlab        = 'Month')
```

We strongly discourage the use of automatic rephasing, as it can lead to misleading interpretations of process stability. By automating shift detection, automatic rephasing effectively removes the essential human insight into the process. Therefore, rephasing should be a deliberate decision made by individuals with a thorough understanding of the process and any associated interventions.

Rephasing *may* be appropriate when the following conditions are met:

* there is a sustained shift in data,
* the reason for the shift is known,
* the shift is in the desired direction, and
* the shift is expected to continue.

If any of these conditions is not met, we should rather seek to understand the nature and causes of any shift, following the strategy outlined in Chapter \@ref(using), [Using SPC in Healthcare].

## Assuming a one-to-one link between PDSA cycles and data points

While this issue may not result in faulty SPC analyses, it often causes confusion among practitioners. SPC is frequently taught alongside improvement work using PDSA cycles, which can lead to the mistaken belief that there must be a one-to-one correspondence between PDSA cycles and data points on an SPC chart. In other words, any detected freak, shift or trend is assumed to be directly linked to specific PDSA cycles testing particular actions.

This is, however, far from always the case. PDSA cycles tend to run quickly (minutes, hours, days), while SPC charts reflect changes more slowly (days, weeks, months).

Therefore, it is important for practitioners to recognise that SPC charts and PDSA cycles operate on different timescales and serve complementary but distinct purposes. SPC should be used as a broader monitoring tool to observe trends and shifts over time, while PDSA cycles focus on rapid, iterative testing and learning. Understanding this distinction helps ensure more effective use of both methods in quality improvement efforts.

## The control chart vs run chart debate

It is a common and persistent misconception that control charts are inherently superior to run charts in detecting special cause variation (@carey2002b). Similarly, run charts and control charts are often viewed as fundamentally different methods, warranting distinct terminologies (@perla2011). We disagree with both notions.

We refer to run charts and control charts collectively as SPC charts -- time series charts that use statistical tests to detect signs of special cause variation. These tests -- known as rules -- employ various techniques to identify patterns in data that are unlikely to arise in stable (i.e., purely random) processes.

All SPC rules share a common basis: they rely on the position of data points relative to the process centre and/or spread. For a detailed discussion of these concepts, see Appendix \@ref(diagnostics), [Basic Statistical Concepts].

Traditionally, control charts use rules based on sigma limits, while run charts apply rules based on the centre line. However, combining rules based on both centre and spread -- as we advocate throughout this book -- is not a new practice. Consider Rule #4 of the Western Electric rules — a run of eight or more consecutive data points on the same side of the centre line. This rule, which is independent of sigma limits, is grounded in the same statistical principle as the rule for unusually long runs.

